---
title: "Microbe and gene correlations"
author: "Rebecca Hoyd"
date: "4/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tidygraph)
library(ggraph)
library(igraph)
library(ggtext)
library(Hmisc)
library(ggalluvial)
library(rlist)
```

# Load data

```{r}
edges.raw <- readRDS("/fs/ess/PAS1695/projects/exotic/data/network_TCC-reg-in.RDS")
exprs.norm <- readRDS("/fs/ess/PAS1695/projects/exotic/data/drake-output/2022-03-08/tcc.expr.norm.RDS")
```

# Filter low variance genes

```{r}
exprs.var <- unlist(lapply(exprs.norm[,-1], var))
table(exprs.var > 1)
genes.keep <- names(subset(exprs.var, exprs.var > 1))
```

# Some quick queries for network characteristics

## Degree centrality

```{r calculate hubbiness}
hub.m <- edges.raw %>%
  filter(gene %in% genes.keep) %>%
  group_by(microbe) %>%
  tally() %>%
  arrange(desc(n)) %>%
  mutate(rank = row_number())

hub.g <- edges.raw %>%
  filter(gene %in% genes.keep) %>%
  group_by(gene) %>%
  tally() %>%
  arrange(desc(n)) %>%
  mutate(rank = row_number())

head(hub.m)
```

```{r}
write.csv(hub.g, "../tables/network_gene_degree-centrality.csv",
          row.names = F)
write.csv(hub.m, "../tables/S2_network_microbe_degree-centrality.csv",
          row.names = F)
```

```{r histogram degree centrality}
degree.cent <- bind_rows(hub.g, hub.m)
write.csv(degree.cent, "../tables/S5_degree-centrality_regression.csv", row.names = F)

degree.cent %>%
  ggplot(aes(x = n)) +
  geom_histogram(bins = 50) +
  labs(x  = "Number of edges per node", y = "Node count") +
  theme_bw() +
  theme(text = element_text(size = 8))
ggsave("../figures/histogram_node-edges-count_reg.pdf", height = 2.5, width = 2.5)
```

## Path centrality

This block of code makes objects that are then used in the scripts network_find-shortest-path.R to calculate betweenness centality before coming back to this document.
```{r format the igraph}
weighted.graph <- edges.raw %>%
  filter(gene %in% genes.keep) %>%
  mutate(new.weight = 1/abs(estimate))
hist(weighted.graph$new.weight)


edges.for.weight <- weighted.graph %>%
  select(gene, microbe)%>%
  as.matrix() %>%
  graph_from_edgelist(directed = F)

weighted.edges <- set.edge.attribute(edges.for.weight,
                                     "weight",
                                     E(edges.for.weight),
                                     weighted.graph$new.weight)

saveRDS(weighted.edges,
        "/fs/ess/PAS1695/projects/exotic/data/network_igraph-data_regression.RDS")

```

```{r format betweenness results}
betweencent <- readRDS("/fs/ess/PAS1695/projects/exotic/data/network_betweenness-centrality_regression.RDS")
betweencent.df <- as.data.frame(betweencent) %>%
  rownames_to_column(var = "node")
write.csv(betweencent.df, "../tables/network_betweenness-centrality_regression.csv",
          row.names = F)
```

```{r get closeness}
closeness.measure <- closeness(weighted.edges, normalized = T)
closeness.form <- closeness.measure %>%
  as.data.frame() %>%
  rownames_to_column(var = "node")
colnames(closeness.form)[2] <- "closeness"
  
closeness.form <- closeness.form %>%
  arrange(desc(closeness))

write.csv(closeness.form, "../tables/network_closeness-centrality_regression.csv",
          row.names = F)
```

# Pathway analysis

## Prepare database

```{r}
library(msigdbr)
library(fgsea)
```

```{r prepare hallmark database}
msig.human <- msigdbr(species = "Homo sapiens") %>%
  dplyr::select(gs_cat, entrez_gene, gs_name, gene_symbol, gs_description) %>%
  dplyr::rename("pathname" = gs_name, "genes" = gene_symbol, "desc" = gs_description,
                "category" = gs_cat, "NCBI Entrez ID" = entrez_gene) 

msig.hlmrk <- msig.human %>%
    filter(category=="H") %>%
    unique()

hlmrk.paths <- msig.hlmrk %>%
     dplyr::select(genes, pathname) %>%
     split(x = .$genes, f = .$pathname)

```

```{r some quick functions}
edgesToFGSEA <- function(m){
  edges.microbe <- edges.raw %>%
    filter(microbe == m) %>%
    filter(gene %in% genes.keep) %>%
    arrange(p.value)
  
  fgsea.ranks <- edges.microbe$estimate
  names(fgsea.ranks) <- edges.microbe$gene
  
  fgsea.res <- fgsea(hlmrk.paths, fgsea.ranks)
  return(fgsea.res)
}

fgseaEnrichPlot <- function(fres, nranks, colname){
  fres %>%
    arrange(pval) %>%
    mutate(rank = row_number(),
           pathway = gsub("HALLMARK_", "", pathway),
           pathway = gsub("_", " ", pathway),
           pathway = capitalize(tolower(pathway)),
           pathway = gsub("Dna", "DNA", pathway),
           pathway = gsub("E2f", "E2F", pathway)) %>%
    filter(rank <= nranks) %>%
    mutate(pathway = fct_reorder(pathway, NES)) %>%
    ggplot(aes(x = NES, y = pathway)) +
    geom_col(fill = colname) +
    labs(x = "Normalized Enrichment", y = "") +
    theme_bw()
}
```

# New figure 3C

```{r}
mics.for.path <- c("s__Pseudomonas sp. SDM007",
                   "s__Alistipes finegoldii", 
                   "s__Bifidobacterium bifidum", 
                   "s__Escherichia coli",
                   "s__Fusobacterium ulcerans")

paths.allmics <- lapply(mics.for.path, edgesToFGSEA)
names(paths.allmics) <- mics.for.path

unpackFSGEAgenes <- function(fgsea.res, m){
  fres.tmp <- fgsea.res %>%
    filter(pval < 0.1)
  glist <- fres.tmp$leadingEdge
  names(glist) <- fres.tmp$pathway
  
  sigpathgenes <- bind_rows(unlist(glist)) %>%
    t() %>%
    as.data.frame() %>%
    rownames_to_column(var = "pathway") %>%
    mutate(pathway = gsub("\\d$", "", pathway),
           microbe = m) %>%
    rename("gene" = "V1")
  
  return(sigpathgenes)
}

pathgenes.long <- lapply(mics.for.path, function(x) try(unpackFSGEAgenes(paths.allmics[[x]], x))) %>%
  list.clean(., is.character) %>%
  bind_rows()
```

## manual network version


```{r, eval = F}
pathgenes.long %>%
  filter(grepl("V2", pathway) | !grepl("\\d$", pathway)) %>%
  mutate(Freq = 1,
         pathway = gsub("_", "\n", gsub("HALLMARK_", "", pathway)),
         microbe = gsub(" ", "\n", gsub("\\w__","", microbe))) %>%
  ggplot(aes(axis1 = microbe, axis2 = gene, axis3 = pathway,
             y = Freq, fill = microbe)) +
  geom_alluvium(show.legend = F) +
  geom_stratum(fill = "white") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)))+
  scale_fill_manual(values = c("red", "darkblue", "darkgreen", "orange", "brown")) +
  theme_void() +
  theme(text = element_text(size = 8))
ggsave("../figures/alluvial_network-pathways_regression.png", width = 10.5)
```

## Facetted Network version

```{r}
pathgene.edgelist <- pathgenes.long %>%
  select(pathway, gene) %>%
  distinct() %>%
  as.matrix()

pathgene.igraph <- igraph::graph_from_edgelist(pathgene.edgelist,
                                               directed = F)

# plot.igraph(pathgene.igraph)

nodecoord <- layout_nicely(pathgene.igraph)
nodelab <- vertex_attr(pathgene.igraph)

nodecoord.form <- nodecoord %>%
  as.data.frame() %>%
  mutate(nodename = nodelab$name)

pathgene.edgelist.anno <- pathgene.edgelist %>%
  as.data.frame() %>%
  rename("nodename" = "pathway") %>%
  left_join(nodecoord.form) %>%
  rename("pathway" = "nodename",
         "path.x" = "V1",
         "path.y" = "V2") %>%
  rename("nodename" = "gene") %>%
  left_join(nodecoord.form) %>%
  rename("gene" = "nodename",
         "gene.x" = "V1",
         "gene.y" = "V2")

pathgene.edgelist.facet.ls <-
  lapply(mics.for.path, function(x){
    pathgene.edgelist.anno %>%
      mutate(microbe = x) %>%
      left_join(mutate(pathgenes.long, include.edge = T))
  } 
  ) %>%
  bind_rows() %>%
  mutate(microbe = gsub("s__", "", microbe),
         colcode = ifelse(include.edge == T, microbe, NA),
         alphcode = ifelse(include.edge == T, "y", "n"))
```

```{r}
ggplot() +
  geom_point(data = nodecoord.form,
             aes(x = V1, y = V2)) +
  geom_segment(data = pathgene.edgelist.facet.ls, 
               aes(x = path.x, xend = gene.x,
                   y = path.y, yend = gene.y,
                   color = colcode),
               alpha = .5,
               show.legend = F) +
  facet_wrap(vars(microbe)) +
  scale_color_manual(values = c("red", "darkblue", "darkgreen", "orange", "brown"),
                     na.value = "grey80") +
  theme_void() +
  theme(text = element_text(size = 8))

ggsave("../figures/network_path-gene_highlight-mic_regression.pdf", height = 3.3, width = 6.3)
```

## LDA style vis

```{r}
paths.effectsize <- lapply(
  names(paths.allmics),
  function(x) paths.allmics[[x]] %>%
    mutate(microbe = x)) %>%
  bind_rows() 

concat.genes <- lapply(paths.effectsize$leadingEdge, function(x) paste(x, collapse = ","))
paths.effectsize$leadingEdge <- unlist(concat.genes)
write.csv(paths.effectsize, "../tables/S8_pathway-enrichment.csv", row.names = F)

LDA.input <- paths.effectsize %>%
  filter(pval < 0.05) %>%
  group_by(microbe) %>%
  arrange(desc(NES)) %>%
  mutate(rank = row_number()) %>%
  filter(rank <= 5)

LDA.input %>%
  filter(microbe %in% c("s__Pseudomonas sp. SDM007", "s__Alistipes finegoldii", "s__Bifidobacterium bifidum")) %>%
  mutate(microbe = gsub("s__", "", microbe),
         pathway = gsub("_", "\n", gsub("HALLMARK_", "", pathway))) %>%
  ggplot(aes(x = pathway, y = NES, fill = microbe)) +
  facet_wrap(vars(microbe), scales = "free_x", nrow = 1) +
  geom_col(show.legend = F) +
  labs(x = "", y = "Normalized effect size") +
  scale_fill_manual(values = c("red", "darkblue", "brown")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 5),
        text = element_text(size = 8))
ggsave("../figures/barplot_LDA-regression-paths.pdf", height = 2.7, width = 6.3)
```
