---
title: "Waterfall plot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(tidyverse)
```

# Access starting targets

```{r}
drakedir <- "/fs/ess/PAS1695/projects/exotic/data/drake-output/2022-03-08/"
# Counts, before any manipulation
raw.tcc.counts <- readRDS(file.path(drakedir, "raw.tcc.counts.RDS")) %>%
  select(-Homo.sapiens)
raw.tcga.counts <- readRDS(file.path(drakedir, "raw.tcga.counts.RDS")) %>%
  select(-Homo.sapiens)


# Counts, removing samples that fail to resolve batches (missing info, etc.)
tcc.batchres <- readRDS(file.path(drakedir, "tcc.batchres.RDS"))
tcga.batchres <- readRDS(file.path(drakedir, "tcga.batchres.RDS"))

tcc.batchres[[1]] <- tcc.batchres[[1]] %>%
  select(-Homo.sapiens)
tcga.batchres[[1]] <- tcga.batchres[[1]] %>%
  select(-Homo.sapiens)

#Counts,after removing contaminants and all microbes not found in TCGA
# loadd(tcc.counts)
# loadd(tcga.counts)
tcc.counts <- readRDS(file.path(drakedir, "tcc.counts.RDS")) %>%
  select(-Homo.sapiens)
tcga.counts <- readRDS(file.path(drakedir, "tcga.counts.RDS")) %>%
  select(-Homo.sapiens)

#Contaminants found in each of the groups
tcc.contams <- readRDS(file.path(drakedir, "tcc.contams.RDS"))
tcga.contams <-readRDS(file.path(drakedir, "tcga.contams.RDS"))



salter.decontam <- read_excel("/fs/ess/PAS1695/exoticpipe/external-data/41586_2020_2095_MOESM2_ESM.xlsx",
                              sheet = 6, skip = 1)
salter.blanks <- read_excel("/fs/ess/PAS1695/exoticpipe/external-data/41586_2020_2095_MOESM2_ESM.xlsx", 
                            sheet = 7, skip = 1)

salter.eval <- bind_rows(salter.decontam, salter.blanks)

salter.eval.rem <- salter.eval %>%
  dplyr::filter(Category == "LIKELY CONTAMINANT")
salter.eval.rem <- unique(salter.eval.rem$Genera) 
 
# loadd(contaminant.details)
# Final counts with taxonomy
# loadd(tcc.exora.taxonomy)
# loadd(tcga.exora.taxonomy)
```

```{r functions}
resolve_contaminants <- function(decontam.contams, filt.counts, threshold){
  # If the decontam result is in salter, I want to use salter's judgment,otherwise the contaminant is removed
  decontam.badmics <- decontam.contams %>%
    dplyr::filter(p < threshold | is.na(p)) %>%
    dplyr::select(microbe, Genera)
  decontam.unq <- decontam.badmics[!duplicated(decontam.badmics[ , c("microbe", "Genera")]),]
  decontam.rem <- subset(decontam.unq, !(decontam.unq$Genera%in%salter.eval$Genera))
  
  approved <- filt.counts %>%
    tidyr::gather(-sample, key = "microbe", value = "counts") %>%
    dplyr::filter(!microbe %in% decontam.rem$microbe) %>%
    tidyr::separate(microbe, into = c("Genera"), remove = F, sep = "\\.") %>%
    dplyr::filter(!Genera %in% salter.eval.rem) %>%
    dplyr::select(-Genera) %>%
    spread(key = "microbe", value = "counts")
  
 return(approved) 
}
```

# 1A - Portion of reads excluded at each step

Steps: start with raw counts. Remove the samples that don't have information required for batches used in decontam, or are in too small batches for decontam to handle. Use decontam to identify contaminant, then adjust by including Poore et. al. blacklist and whitelist information. Contaminants are combined between TCGA and TCC.
```{r}
#Start: sum all raw counts
tcc.startcount = sum(colSums(raw.tcc.counts[, -1]))
tcga.startcount = sum(colSums(raw.tcga.counts[, -1]))

#How many lost when batches fixed?
# tcc.batchcount <- sum(colSums(tcc.batchres$counts[,-1]))
tcc.batchcount <- tcc.startcount
tcga.batchcount <- sum(colSums(tcga.batchres$counts[,-1]))
# Need to split up the contaminant phase, I think

## Remove decontam found contaminants
bad.contams.tcc <- tcc.contams %>%
  filter(contaminant == TRUE)
bad.contams.tcga <- tcga.contams %>%
  filter(contaminant == TRUE)
tcc.rem.decontam <- raw.tcc.counts %>%
  select(-any_of(bad.contams.tcc$microbe))
tcga.rem.decontam <- tcga.batchres$counts %>%
  select(-any_of(bad.contams.tcga$microbe))

tcc.rem.decontam.count <- sum(colSums(tcc.rem.decontam[,-1], na.rm = T))
tcga.rem.decontam.count <- sum(colSums(tcga.rem.decontam[,-1], na.rm = T))

## After the inclusion of the blacklist (additions and subtractions)
tcc.rem.salter <- resolve_contaminants(bind_rows(tcc.contams, tcga.contams), raw.tcc.counts, 0.1)
tcga.rem.salter <- resolve_contaminants(bind_rows(tcc.contams, tcga.contams), tcga.batchres$counts, 0.1)
# tcga.rem.salter <- tcga.batchres$counts %>%
#     tidyr::gather(-sample, key = "microbe", value = "counts") %>%
#     dplyr::filter(!microbe %in% contaminant.details$decontam.remove.salter$microbe) %>%
#     tidyr::separate(microbe, into = c("Genera"), remove = F, sep = "\\.") %>%
#     dplyr::filter(!Genera %in% contaminant.details$salter.likely.contaminants) %>%
#     dplyr::select(-Genera) %>%
#     spread(key = "microbe", value = "counts")

tcc.rem.salter.count <- sum(colSums(tcc.rem.salter[,-1], na.rm = T))
tcga.rem.salter.count <- sum(colSums(tcga.rem.salter[,-1], na.rm = T))

# remove mics from TCC only
tcc.finprep <- tcc.counts
tcc.fincount <- sum(colSums(tcc.finprep[,-1]))
tcga.fincount <- sum(colSums(tcga.counts[,-1]))
```

```{r get waterfall counts into ggplotable df}
waterfall.df <- as.data.frame(cbind(step = c("start", "batch resolution",
                                             "initial decontam",
                                             "literature altered decontam", 
                                             "Remove TCC only"),
                                    tcc = c(tcc.startcount, tcc.batchcount,
                                            tcc.rem.decontam.count,
                                            tcc.rem.salter.count, tcc.fincount),
                                    tcga = c(tcga.startcount, tcga.batchcount,
                                             tcga.rem.decontam.count,
                                             tcga.rem.salter.count, tcga.fincount)))

ymin.tcc <- c(2.8e7, waterfall.df$tcc[1:4])
ymin.tcga <- c(9e8, waterfall.df$tcga[1:4])

waterfall.df <- waterfall.df %>%
  tidyr::gather(-step, key = "Data.Source", value = "count")
waterfall.df$y.min <- c(ymin.tcc, ymin.tcga)
```

## Waterfall format - both datasets
```{r}
# waterfall.df %>%
#   mutate(id = rep(1:5, 2),
#          y.min = as.numeric(y.min),
#          count = as.numeric(count),
#          changedir = ifelse(y.min < count, "Gain", "Loss"),
#          Data.Source = toupper(Data.Source)) %>%
#   ggplot() +
#   geom_rect(aes(ymin = y.min, ymax = count, xmin = id - .45 ,
#             xmax = id + .45, x = id, fill = changedir)) +
#   scale_fill_manual(breaks = c("Gain", "Loss"), values = c("darkgreen", "darkred"), 
#                     name = "Effect on\nread count") +
#   facet_wrap(vars(Data.Source), scales = "free_y") +
#   # scale_x_discrete(breaks = 1:5, labels = c("start", "batch resolution", "initial decontam",
#   #                                            "literature altered decontam", "Remove TCC only")) +
#   labs(x = "") +
#   theme_bw() 
# ggsave("../figures/1A_waterfall-plot.png")
```

## Barplot format

```{r}
# waterfall.df %>%
#   mutate(id = rep(1:5, 2),
#          y.min = as.numeric(y.min),
#          count = as.numeric(count)) %>%
#   ggplot(aes(x = id, y = count)) +
#   geom_col() +
#   facet_wrap(vars(Data.Source), scales = "free_y") +
#   theme_bw()
# ggsave("../figures/fig1_waterfall_barplot.png")
```

## Waterfall format - only TCGA

I think I can combine the two plot types to make something really nice here ...
```{r}
tcga.waterfall.df <- waterfall.df %>%
  mutate(id = rep(1:5, 2),
         y.min = as.numeric(y.min),
         count = as.numeric(count),
         changedir = ifelse(y.min < count, "Gain", "Loss"),
         Data.Source = toupper(Data.Source)) %>%
  filter(id != 5 & Data.Source == "TCGA")
```

```{r}
tcga.waterfall.df %>%
  mutate(id = 6-id,
         y.min = ifelse(id == 5, 0, y.min)) %>%
  ggplot(aes(x = id, y = count)) +
  geom_rect(aes(ymin = 0, ymax = 805457002, xmin = 1 - .45 ,
            xmax = 1 + .45)) +
  geom_rect(aes(ymin = y.min, ymax = count, xmin = id - .45 ,
            xmax = id + .45, fill = changedir)) +
  scale_fill_manual(breaks = c("Gain", "Loss"), values = c("darkgreen", "darkred"), 
                    name = "Effect on\nread count") +
  labs(x = "", y = "") +
  scale_x_continuous(breaks = c(5,4,3,2,1), 
                     labels = c("Initial read count",
                                "Remove samples \nfrom small batches",
                                "Remove contaminants \nfrom decontam",
                                "Adjust contaminants \nwith literature review",
                                "Final read count")) +
  coord_flip() +
  theme_bw() 
ggsave("../figures/waterfall_TCGA-combine-types.pdf")
```

